Parametricity Consequences of GADTs Representations

### 2021-02-18 20-gadts/GADT-story.tex
work on gadt-story

\begin{abstract}
FIX ME!!!! GADTs are misnamed. They don't really generalize ADTs. And
they don't behave the way we expect data types to behave. In
particular, they force us to choose between the naturality properties
we expect of their fmap functions and other parametricity results for
them, causing us to question in what sense GADTs are data types, and
what the value of the parametricity results for them as determined by
their syntax really is.
\end{abstract}

@@

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
equivalent, in the sense that they both have the same interpretation
in the model. Models in which the Church encodings and primitive
representations of all {\color{blue} (polynomial)} algebraic data
types (ADTs) are equivalent include the operationally-based model
of~\cite{pit98,pit00} and the categorical models
of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
This equivalence was recently extended to the class of what might be
called ``hereditarily polynomial'' {\color{blue} to include truly
  nested such types} nested types via the categorical model
of~\cite{jgj21}. But, as shown in Section~\ref{sec:par} below, for
some {\color{blue} (even polynomial)} GADTs,
%that are not hereditarily polynomial nested types,
no such isomorphism holds.

{\color{red} Maybe use a less degenerate example.}  GADTs are
traditionally represented via Church encodings. In this view, GADTs
are completely determined by their syntax. For example, the GADT
$\mathsf{G}$ defined by
\[\begin{array}{l}
\mathsf{data\,G : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}\]
{\color{blue} Use $\top$ instead of $1$ throughout?}

\noindent
{\color{blue} data type vs data structure --- which should have the
functorial completion? The data type.}

\noindent
comprises exactly one element, namely $\mathsf{c}$. By contrast, the
primitive representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. {\color{blue} Completing the
  syntax is necessary to have a {\em functor} to take the least fixed
  point of, so is essential to the view of data types as primitive.}
Viewed as a primitive data type, $\mathsf{G}$ comprises not just
$\mathsf{c}$, but also all data elements of the form
$\mathsf{map\;f\;c}$ for all types $\mathsf{t}$ and all functions
$\mathsf{f : 1 \to t}$ definable in the language, as well as all data
elements of the form $\mathsf{map\;g\;x}$ for each appropriately typed
function $\mathsf{g}$ and each element $\mathsf{x}$ already added to
the data type, and so on.  Although functorial completion adds no new
data elements to the syntax of a (polynomial) ADT or a (hereditarily
polynomial) nested type, so that its primitive Church encoding
representations coincide, these are rarely the same for GADTs.
Indeed, they are quite different for even the very simple GADT
$\mathsf{G}$ above.

@@

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. When used this way, it is more natural to view the GADT in
terms of its primitive representation. In fact, the functorial
completion of a GADT inherent in its primitive representation is
exactly what is needed to ensure that the GADT supports a
$\mathsf{map}$ function, i.e., is a functor both in the catgorical
sense and in the sense of being an instance of Haskell's
$\mathsf{Functor}$ class.  It is worth noting, however, that since the
functorial completion of a GADT includes only those data elements that
are derivable via repeated applications of $\mathsf{map}$ starting
from those specified by its syntax, primitive GADTs need not, in
general, be nested types. For example, the functorial completion of
our GADT $\mathsf{G}$ includes elements at any instance
$\mathsf{G\,t}$ because there is a function from $\mathsf{1}$ to any
type $\mathsf{t}$, but it does not include any terms at instance
$\mathsf{G\,0}$ because no calculus allows the definition of a
function into $\mathsf{0}$. {\color{blue} Use $\bot$ instead of $0$
  here?}

@@

\section{Representations of Data Types}

@@

{\color{blue} polynomial ADTs?} Since the shape of an ADT structure
--- i.e., a structure whose type is an instance of an ADT --- is
independent of the type of data it contains, ADTs can be defined
polymorphically. As a result, an ADT structure containing data of type
$\mathsf{a}$ can be transformed into another ADT structure of the
exact same shape containing data of another type $\mathsf{b}$ simply
by applying a given function $\mathsf{f \,:\,a \to b}$ to each of its
elements. Indeed, every ADT can be made an instance of Haskell's
$\mathsf{Functor}$ class by defining a type-and-data-uniform
type-preserving, data-changing $\mathsf{fmap}$ {\color{blue} map
  vs. fmap} function for it.  Moreover, given a type-independent way
of rearranging an ADT structure's shape $\mathsf{T\,a}$ into the shape
for another ADT structure $\mathsf{T'\,a}$, we get the same structure
of type $\mathsf{T'\,b}$ regardless of whether we first rearrange the
original structure of type $\mathsf{T\,a}$ into one of type
$\mathsf{T'\,a}$ and then use $\mathsf{fmap_{T'}}$ to convert that
resulting structure to one of type $\mathsf{T'\,b}$, or we first use
$\mathsf{fmap_T}$ to convert the original structure of type
$\mathsf{T\,a}$ to one of type $\mathsf{T\,b}$ and then rearrange that
resulting structure into one of type $\mathsf{T'\,b}$. For example, if
$\mathsf{f : a \to b}$, $\mathsf{xs \,:\,List\,a}$, and $\mathsf{g :
  List\,a\to Tree\,a}$ arranges lists into trees in a type-independent
way, then
\[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]

@@

It is possible to make $\mathsf{List}$ and $\mathsf{Tree}$ instances
of Haskell's $\mathsf{Functor}$ class. Indeed, the Prelude's
$\mathsf{map}$ function does exactly that for $\mathsf{List}$. For
$\mathsf{Tree}$ we can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,x\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
But the Church encoding of an ADT carries with it no expectation
whatsoever that such a type-and-data-uniform, type-preserving,
data-changing $\mathsf{map}$ function can be defined. Nevertheless,
for polynomial ADTs this can always be done. But even when the Church
encoding of an ADT supports a $\mathsf{map}$ function, there is no
guarantee that the rearrange-transform property will hold for it.

@@

\subsubsection{Church Encodings of GADTs}

@@

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, a
GADT calculus must support an equality type $\mathsf{Equal}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $\mathit{Equal}$. {\color{blue} This is the least commitment
  thing Chirch encodings require.} Such a type can either be taken as
primitive, as in
\[\begin{array}{l}
\mathsf{data\, Equal :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

@@

\subsubsection{GADTs as Primitive Data Types}

@@

@@

Intuitively it seems the two representations might be different. One
is functorial, one is not. It is possible they could always have the
same data elements even if not the same morphisms. But as we see in
Example~\ref{ex:g1}, this turns out not to be the case.

Non-uniform nature of GADTs renders build, therefore short cut, for
GADTs impossible. Can't even formulate! Of course, there could be
another way to prove equivalence.  But, in fact, CEs and primitive
representations are not equivalent for arbitrary GADTs, as the
following example shows.

\begin{example}\label{ex:g1}
{\color{red} THIS EXAMPLE IS ENRICO'S TEXT}
Consider the following GADT, defined in a Haskell-like syntax:
\begin{equation}\label{eq:example-GADT}
\begin{array}{l}
\mathsf{data }\; G : \ast \to \ast\;\mathsf{where}\\
\hspace*{0.4in}c\;:\;G\;1
\end{array}
\end{equation}
This GADT has a unique constructor, providing a data of type $G\;1$.
While we might say this is the simplest GADT possible, it still
provides an interesting case-study, showing the differences between
the primitive data type representation and the Church encoding
representation.

We now represent $G$ via its Church encoding.  Since $G$ has a single
constructor, which provides a data of type $G\,1$ without requiring
any input, its Church encoding is
\[
G\,\alpha = \mathsf{\forall f.\, (\forall a.\,
  f\,a) \to f\,\onet}
\]
meaning that $G\,A = \mathsf{Equal}\,\onet\,A$.  Thus, $G$ tests its
argument for equality against the unit type, $\onet$.

Recall from the previous section {\color{blue} use reference} that the
semantic interpretation of $\mathsf{Equal}\,\onet\,A$ is given by a
characteristic function, in the sense that it is the singleton 1 when
$A$ is 1, and the empty set otherwise.  Thus, as a data type
represented by Church encoding, G will be the data type containing one
(and only one) piece of data when instantiated on $\onet$, and
uninhabited otherwise.  This perfectly mirrors the syntax of $G$,
which indeed provides just one constructor when the type is
instantiated on 1, and none otherwise.  Thus, given a set $A$, the set
interpretation $G_0(A)$ is 1 if $A = 1$, and $\emptyset$ otherwise.

We now want a primitive representation of $G$.  Recall that $G$ has a
unique constructor $c : G\,\onet$, which can be equivalently expressed
as $c: \onet \to G\,\onet$.  By using a $\Lan$ type, we can express
$c$ as a constructor $(\Lan_{\onet} \onet)\alpha \to G \alpha$.  Thus,
as a primitive data type, we represent $G$ as $\mu \phi. \lambda
\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$.  {\color{blue}
  (WHERE?) notice that the constructors of a GADT have to be functors,
  because the GADT is the fixpoint of a higher order functor.}

The semantic interpretation of $\mu \phi. \lambda
\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$ is given by the
fixpoint of the functor yielded by the semantic interpretation of
$\Lan^{\emptyset}_{\onet} \onet$, the body of the $\mu$ type.  Notice,
though, that the recursive variable $\phi$ does not appear in the body
of the $\mu$ type.  Thus, the semantic interpretation is actually just
the semantic interpretation of the body itself.  Therefore, the
semantic interpretation is $\mathsf{\Lan}_1 1$, where $1$ is the
constantly $1$-valued functor $\ast \to \set$ from the category with a
single object $\ast$ to $\set$.

It turns out that $\mathsf{\Lan}_1 1$ is, in fact, none other than the
identity functor on $\set$.  Indeed, for any set $A$,
$(\mathsf{\Lan}_1 1) A$ is the set of all functions $a : 1 \to A$
modulo the smallest equivalence relation $\sim$ such that, for any $f
: 1 \to 1$, we have $a \sim a \circ f$.  But this is isomorphic to
$A$, because the only function $1 \to 1$ is the identity.
{\color{blue} Reference Computing left Kan extensions - M.R.Bush
  M.Leeming R.F.C.Walters}

@@

1. Parametricity requires a fibred semantics. This doesn't happen for
primitive rep so can't have a par model with this rep. Can still get
naturality properties, however.

2. For CEs can get par model and par properties. But what do these say
for GADTs? Give examples?

{\color{red} EVERYTHING BEFORE THE NOTES BELOW IS ENRICO'S TEXT:} To
even consider the question of whether our semantics is parametric we
first need to define a relational semantics for our types, i.e., a
relational interpretation for every type $G : \ast \to \ast$ such
that, for every relation $R : \rel(A, B)$, we have $G_1(R) :
\rel(G_0(A), G_0(B))$.

@@

Following {\color{blue} Reference Computing left Kan extensions -
  M.R.Bush M.Leeming R.F.C.Walters} we know that
$(\mathsf{\Lan}_{\Eq_1} 1) R$ is the set of all morphisms of relations
$f : \Eq_1 \to R$ modulo an equivalence relation; but there can be no
morphism of relations $\Eq_1 \to R$ because its second component would
have to be a function $1 \to \emptyset$, and there can be no such
function.  Thus, $(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is
not the domain we would have expected.

@@

@@

%%% CE are non functorial but parametric
As we have seen in the previous section {\color{blue} ref}, GADTs
represented as Church encodings do not necessarily have a map or a
functorial semantics.  In particular, this means that we cannot
possibly derive naturality results for those types, event though their
semantics is parametric.  Indeed, even though naturality is sometimes
seen as a consequence of parametricity, that is only true in case we
have a functorial semantics and types supporting a map.  Still, GADTs
represented as Church encodings can still exhibit consequences of
paramentricity other than naturality, as they do admit parametric
semantics {\color{blue} cite Bob's paper?}.

%%% Example with G
{\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
  represent syntax, so should be sans-serif, but in the notes, the
  paragraph about set/relational interpretation and
  parametricity/IEL/AT uses serif font.}  For example, consider the
Church encoding of the GADT $\mathsf{G}$ defined
in~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
$\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
$t_0(A) \in G_0(A)$ and $t_0(B) \in G_0(B)$, and by the abstraction
theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
We also know that $G_1(R)$ is the empty relation whenever $R$ is not
$\mathit{Equal}\,1$.  We thus deduce that there can be no term of type
$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
$\mathsf{1}$.  This result is the syntactic counterpart of the
inhabitation result for $\mathsf{G}$ {\color{blue} are inhabitation
  results about semantics or syntax?}  showing that all of the
instances of $\mathsf{G}$ are uninhabited, with the exception of the
instance for $\mathsf{1}$.  Of course, we could not possibly have a
term of an uninhabited type.

%%% Primitive representation of GADT is non parametric, but functorial
On the contrary, as we have shown previously {\color{blue} ref}, a
GADT represented as a primitive data type does not necessarily admit a
parametric semantics, but, as a fixpoint of a higher-order functor, it
does admit a functorial semantics and, consequently, a functorial map.
In particular, every type that has a functorial map also benefits from
naturality properties.  In other words, there is a suitable
rearrange-transform property as explained in Section~\ref{sec:adts}.

@@

@@

NOTES/POINTS:

Bob requires map functions for fixpoint but then doesn't use them.

>>>>

\begin{abstract}
{\color{red} FIX ME!} GADTs are misnamed. They don't really generalize
ADTs. And they don't behave the way we expect data types to behave. In
particular, they force us to choose between the naturality properties
we expect of their fmap functions and other parametricity results for
them, causing us to question in what sense GADTs are data types, and
what the value of the parametricity results for them as determined by
their syntax really is.
\end{abstract}

@@

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
semantically equivalent, i.e., both have the same interpretation in
the model. Models in which the Church encodings and primitive
representations of all {\color{blue} (polynomial)} algebraic data
types (ADTs) are semantically equivalent include the
operationally-based model of~\cite{pit98,pit00} and the categorical
models of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
Correctness of short cut fusion, and thus semantic equivalence of
Church encodings and primitive representations, was recently extended
to the class of what might be called ``hereditarily polynomial''
{\color{blue} to include truly nested such types} nested types via the
categorical model of~\cite{jgj21}. But, as shown in
Section~\ref{sec:par} below, for some {\color{blue} (even polynomial)}
GADTs,
%that are not hereditarily polynomial nested types,
no such isomorphism holds.

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax; for example,
the GADT $\mathsf{G}$ defined in Haskell by
\begin{equation}\label{eq:example-GADT}
\begin{array}{l}
\mathsf{data\, Expr\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Expr\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Expr \,a \to Expr\,b \to
  Expr\,(a \times b)}\\ 
\end{array}
\end{equation}
%\[\begin{array}{l}
%\mathsf{data\,G : * \to *\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
%\end{array}\]
%{\color{blue} Use $\top$ instead of $1$ throughout?}
\noindent
comprises expressions of any type $\mathsf{a}$ and expressions
obtained by pairing other expressions. By contrast, the primitive
representation sees a GADT as the {\em functorial
  completion}~\cite{jp19} of its syntax. Completing a GADT's syntax to
a {\em functor} is necessary in order to be able to take the least
fixed point~\cite{tfca}, so it is inherent in the view of data types
as their primitive representations.  Viewed as its primitive
representation, $\mathsf{Expr}$ comprises not just those data elements
representable in syntax, but also all data elements of the form
$\mathsf{map\,f\,e}$ for all types $\mathsf{t_1}$ and $\mathsf{t_1}$,
all functions $\mathsf{f : t_1 \to t_2}$ definable in the language,
and all $\mathsf{e :: Expr\,t_1}$, as well as all data elements of the
form $\mathsf{map\,g\,e'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{e'}$ already added to the data
type, and so on. Functorial completion adds no new data elements to
the syntax of a GADT that is actually a (polynomial) ADT or a
(hereditarily polynomial) nested type, so its Church encoding and
primitive representation coincide. However, these representations are
rarely the same for GADTs.
%Indeed, they are quite different for even the very simple GADT
%$\mathsf{G}$ above.

@@

A person interested in the semantics of functional languages, on the
other hand, is likely to follow~\cite{bfss90} and expect GADTs to
behave as their names suggests --- i.e., as generalizations of
ADTs. The semanticist is thus likely to expect a GADT to be able to be
used as a ADT is used --- i.e., as a shape filled with data that can
be changed without also changing the shape. When used this way, it is
more natural to view the GADT in terms of its primitive
representation. In fact, the functorial completion of a GADT inherent
in its primitive representation is exactly what is needed to ensure
that the GADT supports a $\mathsf{map}$ function, i.e., is a functor
both in the catgorical sense and in the sense of being an instance of
Haskell's $\mathsf{Functor}$ class.  It is worth noting, however, that
since the functorial completion of a GADT includes only those data
elements that are derivable via repeated applications of
$\mathsf{map}$ starting from those specified by its syntax, primitive
GADTs need not, in general, be nested types. For example, the
functorial completion of our GADT $\mathsf{G}$ includes elements at
any instance $\mathsf{G\,t}$ because there is a function from
$\mathsf{1}$ to any type $\mathsf{t}$, but it does not include any
terms at instance $\mathsf{G\,0}$ because no calculus allows the
definition of a function into $\mathsf{0}$. {\color{blue} Use $\bot$
  instead of $0$ here?}

@@

\section{Representations of Data Types}\label{sec:reps}

@@

{\color{blue} polynomial ADTs?} Since the shape of an ADT structure
--- i.e., a structure whose type is an instance of an ADT --- is
independent of the type of data it contains, ADTs can be defined
polymorphically. As a result, an ADT structure containing data of type
$\mathsf{a}$ can be transformed into another ADT structure of the
exact same shape containing data of another type $\mathsf{b}$ simply
by applying a given function $\mathsf{f \,:\,a \to b}$ to each of its
elements. Indeed, every ADT can be made an instance of Haskell's
$\mathsf{Functor}$ class by defining a type-and-data-uniform
type-preserving, data-changing $\mathsf{fmap}$ {\color{blue} map
  vs. fmap} function for it.  Moreover, given a type-independent way
of rearranging an ADT structure's shape $\mathsf{T\,a}$ into the shape
for another ADT structure $\mathsf{T'\,a}$, we get the same structure
of type $\mathsf{T'\,b}$ regardless of whether we first rearrange the
original structure of type $\mathsf{T\,a}$ into one of type
$\mathsf{T'\,a}$ and then use $\mathsf{fmap_{T'}}$ to convert that
resulting structure to one of type $\mathsf{T'\,b}$, or we first use
$\mathsf{fmap_T}$ to convert the original structure of type
$\mathsf{T\,a}$ to one of type $\mathsf{T\,b}$ and then rearrange that
resulting structure into one of type $\mathsf{T'\,b}$. For example, if
$\mathsf{f : a \to b}$, $\mathsf{xs \,:\,List\,a}$, and $\mathsf{g :
  List\,a\to Tree\,a}$ arranges lists into trees in a type-independent
way, then
\[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]
{\color{violet} mention naturality?}

@@

It is possible to make $\mathsf{List}$ and $\mathsf{Tree}$ instances
of Haskell's $\mathsf{Functor}$ class. Indeed, the Prelude's
$\mathsf{map}$ function does exactly that for $\mathsf{List}$. For
$\mathsf{Tree}$ we can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,x\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
But the Church encoding of an ADT carries with it no expectation
whatsoever that such a type-and-data-uniform, type-preserving,
data-changing $\mathsf{map}$ function can be defined. Nevertheless,
for polynomial ADTs this can always be done. But even when the Church
encoding of an ADT supports function with the same type as its actual
$\mathsf{map}$, there is no guarantee that the rearrange-transform
property will hold for it. For example, if {\color{violet} Are there
  examples of such non-functorial maps?}

@@

\subsubsection{Church Encodings of GADTs}\label{sec:CEs}

@@

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, a
GADT calculus must support an equality type $\mathsf{Equal}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $\mathit{Equal}$. {\color{blue} This is the least commitment
  thing Church encodings require.} Such a type can either be taken as
primitive, as in
\[\begin{array}{l}
\mathsf{data\, Equal :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

@@

\subsubsection{GADTs as Primitive Data Types}\label{sec:prim-gadts}

@@

{\color{blue} rearrange-transform property for GADTs?}

@@

{\color{red} Check all fonts!! Syntax vs. semantics!}  Intuitively it
seems the two representations might be different. One is functorial,
one is not. It is possible they could always have the same data
elements even if not the same morphisms. But as we will see in
Example~\ref{ex:g1}, this turns out not to be the case.

Non-uniform nature of GADTs renders build, therefore short cut, for
GADTs impossible. Can't even formulate! Of course, there could be
another way to prove equivalence.  But, in fact, CEs and primitive
representations are not equivalent for arbitrary GADTs, as the
following example shows.
{\color{blue} Our favorite GADT isn't a hereditarily
  polynomial nested type.}

\begin{example}\label{ex:g1}
Consider the following GADT, defined in a Haskell-like syntax:
\begin{equation}\label{eq:example-GADT}
\begin{array}{l}
\mathsf{data\,G : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}
\end{equation}
Syntactically, this GADT comprises a single data element, namely
$\mathsf{c :: G \, 1}$. Despite the simplicity of this GADT, it serves
as an informative case study highlighting the differences between the
Church encoding of a GADT and that GADT's representation as a
primitive data type.

The Church encoding of $\mathsf{G}$ is
\[\mathsf{G\,a} = \mathsf{\forall f.\, (\forall b.\,
  f\,b) \to f\,1} \] In other words, $\mathsf{G\,a =
  \mathsf{Equal}\,1\,a}$, so $\mathsf{G}$'s effect is simply to test
its argument for equality against the unit type $\mathsf{1}$.

Recall from the Section~\ref{sec:CEs} that the semantic interpretation
of $\mathsf{Equal\,1\,a}$ is given by a characteristic function, in
the sense that it is the singleton $1$ when the interpretation of
$\mathsf{a}$ is $1$, and is the empty set otherwise.  Thus, as a data
type represented by Church encoding, $\mathsf{G}$ will be the data
type containing one (and only one) piece of data when instantiated on
$\mathsf{1}$, and uninhabited otherwise.  This perfectly mirrors the
syntax of $\mathsf{G}$, which indeed provides just one constructor
when the type is instantiated on $\mathsf{1}$, and none otherwise.
Thus, given a set $A$ interpreting $\mathsf{a}$, the set
interpretation $G_0(A)$ is $1$ if $A = 1$ and $\emptyset$ otherwise.

We now want a primitive representation of $\mathsf{G}$.  Recall that
$\mathsf{G}$ has a unique constructor $\mathsf{c : G\,1}$, which can
be equivalently expressed as $\mathsf{c: 1 \to G\,1}$.  By using a
a$\Lan$ type, we can express $\mathsf{c}$ as a constructor
$\mathsf{(Lan_1\, 1)\alpha \to G \alpha}$.  Thus, as a primitive data
type, we represent $\mathsf{G}$ as $\mu \phi. \lambda
\alpha. (\Lan^{\emptyset}_1 \,1)\alpha$.  {\color{blue} (WHERE?)
  Notice that the constructors of a GADT have to be functors, because
  the GADT is the fixpoint of a higher order functor. Also, we're not
  using the superscript notation on $\Lan$.}

The semantic interpretation of $\mathsf{\mu \phi. \lambda
  \alpha. (\Lan^{\emptyset}_1\, 1)\alpha}$ is given by the fixpoint of
the functor yielded by the semantic interpretation of
$\mathsf{Lan^{\emptyset}_1\, 1}$, the body of the $\mathsf{\mu}$-type.
Notice, though, that the recursive variable $\mathsf{\phi}$ does not
appear in the body of the $\mathsf{\mu}$-type.  Thus, the semantic
interpretation is actually just the semantic interpretation of the
body itself.  Therefore, the semantic interpretation is
$\mathsf{Lan_1 1}$, where $1$ is the constantly $1$-valued functor
$\ast \to \set$ from the category with a single object $\ast$ to
$\set$.

{\color{red} Syntax? Semantics? Which are we talking here?}
It turns out that $\mathsf{\Lan}_1 1$ is, in fact, none other than the
identity functor on $\set$.  Indeed, for any set $A$,
$(\mathsf{\Lan}_1 1) A$ is the set of all functions $a : 1 \to A$
modulo the smallest equivalence relation $\sim$ such that, for any $f
: 1 \to 1$, we have $a \sim a \circ f$.  But this is isomorphic to
$A$, because the only function $1 \to 1$ is the identity~\cite{blw03}.

@@

\begin{example}
{\color{red} PRIMITIVE REP DOESN'T HAVE FIBRED SEMANTICS, AS REQUIRED
  FOR PAR} To
even consider the question of whether our semantics is parametric we
first need to define a relational semantics for our types, i.e., a
relational interpretation for every type $G : \ast \to \ast$ such
that, for every relation $R : \rel(A, B)$, we have $G_1(R) :
\rel(G_0(A), G_0(B))$.

@@

Following~\cite{blw03} we know that $(\mathsf{\Lan}_{\Eq_1} 1) R$ is
the set of all morphisms of relations $f : \Eq_1 \to R$ modulo an
equivalence relation; but there can be no morphism of relations $\Eq_1
\to R$ because its second component would have to be a function $1 \to
\emptyset$, and there can be no such function.  Thus,
$(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is not the domain we
would have expected.
\end{example}

@@

We still get naturality properties, though.

@@

As we have seen in the previous Section~\ref{sec:reps}, GADTs
represented as Church encodings do not necessarily have a map or a
functorial semantics.  In particular, this means that we cannot
possibly derive naturality results for those types, event though their
semantics is parametric.  Indeed, even though naturality is sometimes
seen as a consequence of parametricity, that is only true in case we
have a functorial semantics and types supporting a map.  Still, GADTs
represented as Church encodings can still exhibit consequences of
paramentricity other than naturality, as they do admit parametric
semantics~\cite{atk12}.

\begin{example}
{\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
  represent syntax, so should be sans-serif, but in the notes, the
  paragraph about set/relational interpretation and
  parametricity/IEL/AT uses serif font.} For example, consider the
Church encoding of the GADT $\mathsf{G}$ defined in
Equation~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
$\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
$t_0(A) \in G_0(A)$ and $t_0(B) \in G_0(B)$, and by the abstraction
theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
{\color{blue} Here, $G = (G_0,G_1)$ is the interpretation of
  $\mathsf{G}$.}  We also know that $G_1(R)$ is the empty relation
whenever $R$ is not $\mathit{Equal}\,1$.  We thus deduce that there
can be no term of type $\mathsf{G\,a}$ unless $\mathsf{a}$ is
semantically equivalent to $\mathsf{1}$.  This result is the syntactic
counterpart of the inhabitation result for $\mathsf{G}$ {\color{blue}
  are inhabitation results about semantics or syntax?}  showing that
all of the instances of $\mathsf{G}$ are uninhabited, with the
exception of the instance for $\mathsf{1}$.  Of course, we could not
possibly have a term of an uninhabited type.
\end{example}

%%% Primitive representation of GADT is non parametric, but functorial
On the contrary, as we have shown in Section~\ref{sec:gadts}, a
GADT represented as a primitive data type does not necessarily admit a
parametric semantics, but, as a fixpoint of a higher-order functor, it
does admit a functorial semantics and, consequently, a functorial map.
In particular, every type that has a functorial map also benefits from
naturality properties.  In other words, there is a suitable
rearrange-transform property as explained in Section~\ref{sec:gadts}.

@@

We can read our counterexample as showing that parametricity results
for calculi representing GADTs as their Church encodings cannot be
extended to calculi that represent GADTs as primitive data types. And
functoriality results for calucli representing GADTs as primitive
cannot be extended to calculi that represent GADTs as their Church
encodings.  So the user is forced to make a choice between naturality
consequences of parametricity and other consequences of
parametricity. If GADTs really did generalize ADTs, and really were
data types in a normal sense, then these would coincide and the user
would not have to choose.

\begin{comment}
{\color{violet} This is already in the paper, split in multiple places.}
An analysis of the Eq A B type in Bob's systems shows that the set
interpretation is empty whenever the A and B have different set
interpretations.  Likewise, its relational interpretation is the empty
relation whenever the A and B have different relational
interpretations (even when the underlying set interpretations are the
same).  The consequences of parametricity are thus immediate: there
can be no term of type Eq A B if A and B are different types.
Moreover, this behavior is fundamentally non-functorial, so that there
is no hope of defining a map for Eq.  In particular, that means we
cannot expect to retrieve non-trivial naturality properties for the Eq
type, despite these are usually considered to be a special case of
parametricity properties.  Indeed, the common understanding that
parametricity generalizes naturality only makes sense for functorial
types.  We can specialize these considerations to the type Eq 1 B,
representing G, our example GADT defined by a unique constructor c ::
G 1.
\end{comment}

@@

NOTES:

{\color{violet} TRUE????}
FOR ADTs and Nesteds, it does not depend. We get the same par property
whether they are seen as primitives or we use their CE.

\subsection{Comparison with Bob's Parametric Model}

Moreover, the consequences of parametricity that one has in Bob's
system do not hold in our system, as we can construct terms of type
$(Lan_1 1)$ B for instances of B other than 1.  Thus, either our
system does not admit any parametric model, or, if it does, the
consequences of parametricity are different.  That the consequences of
parametricity depended on the model or the representation of GADTs is
to be expected, however. (IT IS. CONSIDER INTERPRETING BASE TYPES AS
ORDERED SETS. THIS GIVES DIFFERENT PARAMETRICITY PROPERTIES THAN
INTERPRETING THEM AS DIAGONAL RELATIONS.)
